---
title: "Practical Machine Learning Course Project Write Up"
author: "Anne Domnic"
date: "June 19, 2015"
output: html_document
---

#Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)

#Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

#Project Goal

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

#Reproducibility

An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results below (please see code below for ways and syntax to do so).

#Required Packages

Installing packages and loading required libraries

```{r}
#install.packages("caret")
#install.packages("randomForest")
#install.packages("rpart")
library(caret)
```

```{r}
#Random forest for classification and regression
library(randomForest)
```

```{r}
# Regressive partitioning and regression trees
library(rpart)
# Decision Tree Plot
library(rpart.plot) 

# setting the overall seed for reproduceability
set.seed(1234)
```

#Load the data

The trining data set can be found in the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```

The testing data set can be found in the following URL:

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Load the data to only in memory
```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

#Partitioning the training data set to allow cross -validation

In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (60%) and subTest (40%).

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining) 
dim(myTesting)
```

#Cleaning the data

Three different transformations were used to clean the data as follows:
#1
Cleaning NearZeroVariance Variables Run this code to view possible NZV Variables:

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```

Run this code to create another subset without NZV variables.

```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]

#To check the new set of observations
dim(myTraining)
```

#2
Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:

```{r}
myTraining <- myTraining[c(-1)]
```

#3
Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA's I'm going to leave them out:

```{r}
#creating another subset to iterate in loop
trainingV3 <- myTraining
# every colum in the training dataset
for(i in 1:length(myTraining)) { 
  #if n? NAS > 60% of total observations
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { 
        for(j in 1:length(trainingV3)) {
  #if the columns are the same
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  {
  #remove that column
                trainingV3 <- trainingV3[ , -j] 
            }   
        } 
    }
}
#To check the new N? of observations
dim(trainingV3)
```

```{r}
#Seting back to our data set
myTraining <- trainingV3
rm(trainingV3)
```

Now let us use 3 transformation to our testing data

```{r}
clean1 <- colnames(myTraining)
#"classe" column removed
clean2 <- colnames(myTraining[, -58]) 
myTesting <- myTesting[clean1]
testing <- testing[clean2]

#To check the new N? of observations
dim(myTesting)
```

```{r}
#To check the new N? of observations
dim(testing)
```

#Decision Tree using ML Algorithms for prediction:

In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
#To make sure Coertion really worked, simple smart ass technique:
testing <- rbind(myTraining[2, -58] , testing) 
testing <- testing[-1,]
```

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")

library(rattle)
#To view the dicision tree with fancy, need to run this code
fancyRpartPlot(modFitA1)
```

Predicting
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

Using confusion Matrix to test results
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```

#Using Random Forest for ML Algorithm for Prediction
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

Predicting in-sample error:

```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```

Using confusion Matrix to test results:

```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```

#Conclusion

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.9982 (95% CI: (0.997, 0.999)) compared to 0.8741 (95% CI: (0.8665, 0.8813)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.9982.Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

#Submission

```{r}
#predict outcome levels on the original Testing data set using Random Forest algorithm
predictionsB2 <- predict(modFitB1, testing, type = "class")
predictionsB2
```

```{r}
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```